### 2021-09-28 papers/thesis/resubmitted/ch2-spherical-harmonics/spherical-harmonics.tex
Fixed further typos and corrections from AT notes

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the spherical harmonics as polynomials in three variables $x, y, z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain \nomenclature[Omega]{$\Omega$}{The domain of interest}
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2}
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why this is a practical way to define our domain.

It is useful to be able to transform between these cartesian coordinates and the spherical coordinates \nomenclature[phi]{$\varphi$}{The polar angle in spherical coordinates} \nomenclature[theta]{$\theta$}{The azimuthal angle in spherical coordinates}($\varphi, \theta$). Throughout this thesis we will use the convention that the spherical coordinate angles be defined by \nomenclature[xyz]{$x,y,z$}{Lower case non-bold $x,y,z$ represents cartesian coordinates}
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi
}
where $\rho(z) := \sqrt{1-z^2}$.

In this chapter, we demonstrate how one can write the spherical harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x,y,z$. We demonstrate how these in turn lead to an algorithm for evaluating the spherical harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families of other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere. These examples consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator): written in spherical coordinates this is \nomenclature[Deltas]{$\DeltaS$}{The Laplace--Beltrami operator}
\bseqnnumber{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo \label{eqn:laplacebeltrami}
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for some function $f$ where $\xvec := (x,y,z) \in \R^3$.

@@

Before we proceed further, let us build up to our definition of the spherical harmonics. We first need to introduce a few classical orthogonal polynomials. Orthogonal polynomials are defined as such by the condition that each polynomial is orthogonal to each other, with respect to a given inner product. In particular, they are orthogonal with respect to all lower degree polynomials.

On the unit interval, $[-1,1]$, we note that there is a hierarchy of OPs in the sense that \cite[table 18.3.1, eqn 18.9.15]{DLMF}: \nomenclature[Pnab]{$P^{(a,b)}_n$}{The degree $n$ Jacobi polynomial (context determines normalized or not) unless otherwise stated}
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x) \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x)
}
where $P^{(a,b)}_l (x)$ is the $l$ degree \textit{Jacobi polynomial}, and \nomenclature[Pl]{$P_l$}{The degree $l$ Legendre polynomial} $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$ \nomenclature[Kronecker delta]{$\delta_{n, n'}$}{The Kronecker delta function; $1$ if $n=n'$, else $0$}
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)} \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'}.
}
Legendre polynomials are special cases of \textit{Legendre functions} when $l$ is a non-negative integer. Legendre functions are a class of univariate functions that are solutions to the differential equation
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x} y + l(l+1)y = 0
}
which is helpfully known as the \textit{Legendre equation} \cite[14.2.1]{DLMF}. Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Legendre polynomial \cite[p5]{belousov1962tables}: \nomenclature[Plm]{$\Plm$}{The Associated Legendre polynomial of degree $l$ and order $m$}
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x)
}
for $m = -l, \dots, -1$.

Once again, associated Legendre polynomials are special cases of \textit{associated Legendre functions} when $l$ is a non-negative integer and $m \in \{0,\dots,l\}$, which are also a class of univariate functions that are solutions to a modified version of the Legendre equation
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x}y + \Big[l(l+1) - {m^2 \over 1-x^2} \Big] y = 0
}
which is also helpfully known as the \textit{associated Legendre equation} (\cite[p2]{belousov1962tables}, \cite[14.2.2]{DLMF}). We can see that Legendre functions (and polynomials) are just special cases of the associated versions when $m = 0$.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}: \nomenclature[Ylm]{$\Ylm$}{The degree $l$ and order $m$ spherical harmonic}
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where \nomenclature[clm]{$\clm$}{The normalising constant for the spherical harmonic $\Ylm$}
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}.
}
The spherical harmonics as defined are then orthonormal with respect to the inner product with uniform measure \nomenclature[conjuagate]{$a^*$}{The complex conjugate of $a \in \C$}
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of \nomenclature[C]{$\C$}{The complex numbers} $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any \nomenclature[Z]{$\Z$}{The integers} $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in \nomenclature[R]{$\R$}{The real numbers} \nomenclature[Rd]{$\R^d$}{The d-dimensional coordinate space over the real numbers} $\R^3$.

@@

Jacobi operators are matrices that correspond to multiplication of the orthogonal polynomial basis, in this case the spherical harmonics, by our cartesian coordinates $x, y, z$. We can obtain the entries to these operator matrices by determining the coefficients of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$ in terms of $Y^{m'}_{l'}(x,y,z)$ for any point $(x,y,z)$ on the unit sphere. Relations concerning the products of spherical harmonics have already been established, and so the expressions we desire here could just be seen as special cases of these. More specifically, the coefficients of the expansion in the SH basis of the product of two SHs are composed of Clebsch-Gordan coefficients \cite[p231]{sakurai1995modern}. Clebsch-Gordan coefficients are important in quantum mechanics and angular momentum, with tables and algorithms existing to calculate them \cite{mcnamee1964tables}. By simply choosing appropriate combinations, one could gain the desired coefficients for the expansion of each of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$.

@@

	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[18.9.5, 18.9.6]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is simply a rearranging of the established three-term recurrence detailed in \cite[18.9.1, 18.9.2]{DLMF}

@@

\remark The results in \bsreflemma{lemma:shxyzrelations} can be derived via the relationships established for products of spherical harmonics and Clebsch-Gordan coefficients as mentioned earlier \cite[p231]{sakurai1995modern} but we include this proof so as to generalise to other OPs later on in later chapters.

@@

These recurrences lead to Jacobi operators that correspond to multiplication by $x,y,z$ of the OPs. Define for $l \in \No$: \nomenclature[Pvector]{$\SHvec$}{Vector of the SH OPs} \nomenclature[Pvectorl]{$\SHvec_l$}{(Sub-)Vector of the degree $l$ SH OPs}
\bseqn{
	\SHvecl := 
		\begin{pmatrix}
			Y^{-l}_l \\
			\vdots \\
			Y^l_l
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\SHvec := 
		\begin{pmatrix}
			\SHvec_0 \\
			\hline
			\SHvec_1 \\
			\hline
			\SHvec_2 \\
			\vdots \\
		\end{pmatrix}.
}
Note that these OP vectors we have defined are grouped and ordered according to the degree $l$, and within each degree ordered by the Fourier mode $m$. Let the Jacobi matrices \nomenclature[J]{$J_x, J_y, J_z$}{Jacobi operator matrices} $J_x, J_y, J_z$ hence be given by 
\bseqnnumber{
	J_x \SHvec(x,y,z) = x \SHvec(x,y,z), \quad J_y \SHvec(x,y,z) = y \SHvec(x,y,z), \quad J_z \SHvec(x,y,z) = z \SHvec(x,y,z) \label{eqn:jacobimatsdef}
}
for $(x,y,z) \in \Omega$. While $J_x, J_y, J_z$ are not Jacobi matrices in the classical sense, they are block Jacobi matrices. However, it is useful to simply label them as such. These Jacobi matrices can act on the coefficients vector of a function's expansion in the spherical harmonic basis to implement the operation for multiplication by $x$, $y$ or $z$. For example, given sufficiently large $n \in \N$ let the function $f(x,y,z) : \Omega \to \C$ be given by its expansion \nomenclature[fvec]{$\vec{f}, \vec{\xi}$}{Lower case bold letters refer to vectors} \nomenclature[transpose]{$\vec{v}^\top, A^\top$}{Transpose of the vector $\vec{v}$ or matrix $A$ (non complex conjuagate)}
\bseqn{
	f(x,y,z) = \sum_{l=0}^{N} \sum_{m=-l}^{l} f_{l,m} \; \Ylm(x,y,z) = \SHvec(x,y,z)^\top \: \fvec
}
where 
\bseqn{
	\fvec_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			\vdots \\
			f_{l,-l}
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\fvec := 
		\begin{pmatrix}
			\fvec_0 \\
			\vdots \\
			\fvec_N
		\end{pmatrix}.
}
We say that $\fvec$ is the \textit{coefficients vector} of the function $f$. Then $x f(x,y,z)$ is approximated by $\SHvec(x,y,z)^\top \: J_x^\top \: \fvec$. In other words, $J_x^\top \: \fvec$ is the coefficients vector for the expansion of the function $(x,y,z) \mapsto x \: f(x,y,z)$ in the spherical harmonics basis.

@@

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jz}}
		\centering
		\caption{$J_z$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of the Jacobi operator matrices, showing their sparsity and structure, up to order $N=10$. Each are block-tridiagonal, with the blocks corresponding to polynomial degree.}
	\label{fig:sh-spy-jacobi}
\end{figure}

@@

In spectral space, we wish to work only with vectors of coefficients for the expansion of a function, to which we can apply operator matrices to that represent differential or other operations. Naturally, we of course need a way to obtain the coefficients $f_{l,m}$. The coefficients can be calculated via the integral
\bseqn{
	f_{l,m} = \int_\Omega f(\costheta \sinphi, \sintheta \sinphi, \cosphi) \: \Ylmfull^* \: \sinphi \: \D \varphi \: \D \theta
}
using the orthonormality of the spherical harmonics. As discussed in \bsrefchapter{CHAPTER:intro}, methods to calculate these coefficients exist known as spectral transforms, and are well established for the spherical harmonics (see e.g. \cite{mohlenkamp1999fast, suda2002fast}). Additionally, packages are available for public use that implement efficient spherical harmonics transforms, for example the FastTransforms package \cite{slevinsky2019fast}.

@@

Let $\vec{v}$ be the coefficients of the expansion of $v$ up to order $N$. The operator $V$ would then be the result of the following \textit{operator Clenshaw algorithm}:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+1} = \vec{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \vec{v}_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(J_x,J_y,J_z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } V = V(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0
}
where at each iteration, $\vec{\xi}_n$ is a vector of matrices (note that we are abusing notation here a bit, however it is the simplest way to present the algorithm without introducing yet more matrices!).

>>>>

To introduce ourselves to the world of multidimensional orthogonal polynomials for solving PDEs on the sphere, we can naturally choose to look at the famous spherical harmonics. Our aim here is to express the spherical harmonics as polynomials in three variables $x$, $y$, and $z$ to evaluate functions and solve PDEs on the whole sphere. More precisely, we desire the solution to partial differential equations on the domain \nomenclature[Omega]{$\Omega$}{The domain of interest}
\bseqn{
	\Omega := \bigS^2 = \bsset{(x,y,z) \in \R^3}{x^2 + y^2 = \rho(z)^2}
}
where 
\bseqn{
	\rho(z) := \sqrt{1-z^2}.
}
While it may seem somewhat odd, it should hopefully become apparent why using $\rho$ here is a practical way to define our domain.

It is useful to be able to transform between these cartesian coordinates and the spherical coordinates \nomenclature[phi]{$\varphi$}{The polar angle in spherical coordinates} \nomenclature[theta]{$\theta$}{The azimuthal angle in spherical coordinates}($\varphi, \theta$). Throughout this thesis we will use the convention that the spherical coordinate angles be defined by \nomenclature[xyz]{$x,y,z$}{Lower case non-bold $x$, $y$, and $z$ represent cartesian coordinates}
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta, \\
	y &= \sinphi \sintheta = \rho(z) \sintheta, \\
	z &= \cosphi
}
where $\rho(z) := \sqrt{1-z^2}$.

In this chapter, we demonstrate how one can write the spherical harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x$, $y$, and $z$. We demonstrate how these in turn lead to an algorithm for evaluating the spherical harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families on other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere. These examples consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator): written in spherical coordinates this is \nomenclature[Deltas]{$\DeltaS$}{The Laplace--Beltrami operator}
\bseqnnumber{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo \label{eqn:laplacebeltrami}
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for some function $f$ where $\xvec := (x,y,z) \in \R^3$.

@@

Before we proceed further, let us build up to our definition of the spherical harmonics. We first need to introduce a few classical orthogonal polynomials. Orthogonal polynomials are defined as such by the condition that each polynomial is orthogonal to each other, with respect to a given inner product, and ordered by degree. In particular, they are orthogonal with respect to all lower degree polynomials.

On the unit interval, $[-1,1]$, we note that there is a hierarchy of OPs in the sense that \cite[Table 18.3.1, (18.9.15)]{DLMF}: \nomenclature[Pnab]{$P^{(a,b)}_n$}{The degree $n$ Jacobi polynomial (context determines normalized or not) unless otherwise stated}
\bseqn{
	& \ddx{x} P^{(a,b)}_l (x) = \half \: (l + a + b + 1) \: P^{(a+1,b+1)}_{l-1}(x) \\
	\implies & \dmdxm{x}{m} P_l(x) = {(l+m)! \over 2^m \: l!} \: P^{(m,m)}_{l-m}(x)
}
where $P^{(a,b)}_l (x)$ is the degree $l$ \textit{Jacobi polynomial} with $a, b > -1$, and \nomenclature[Pl]{$P_l$}{The degree $l$ Legendre polynomial} $P_l(x) := P^{(0,0)}_l (x)$ is simply the \textit{Legendre polynomial} of degree $l$. Jacobi polynomials are orthogonal with respect to the weight $w(x) = (1-x)^a(1+x)^b$; that is for $l, l' \in \No,$ \nomenclature[Kronecker delta]{$\delta_{n, n'}$}{The Kronecker delta function; $1$ if $n=n'$, else $0$}
\bseqn{
	\int_{-1}^1 {P^{(a,b)}_l(x)}^2 \:(1-x)^a(1+x)^b \: \D x &=: \omega_{P,l}^{(a,b)}, \\
	\int_{-1}^1 P^{(a,b)}_l(x) \: P^{(a,b)}_{l'}(x) \:(1-x)^a(1+x)^b \: \D x &= \omega_{P,l}^{(a,b)} \: \delta_{l, l'},
}
where $\delta_{l, l'}$ is the regular Kronecker delta function. For the Jacobi polynomials,
\bseqn{
	 \omega_{P,l}^{(a,b)} = {2^{a + b + 1} \; \Gamma(l + a + 1) \; \Gamma(l + b + 1) \over (2l + a + b + 1) \; \Gamma(l + a + b + 1) \; \Gamma(l + 1)}
}
where $\Gamma(x)$ is the regular Gamma function \cite[(5.2.1)]{DLMF} (with $\Gamma(n) = (n-1)!$ for $n \in \N$). Legendre polynomials are special cases of \textit{Legendre functions} when $l$ is a non-negative integer. Legendre functions are a class of univariate functions that are one set of solutions to the differential equation
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x} y + l(l+1)y = 0, \quad x \in [-1,1]
}
which is helpfully known as the \textit{Legendre equation} \cite[(14.2.1)]{DLMF}. Further, the \textit{associated Legendre polynomials} are a set of polynomials orthogonal with respect to unit weight on the unit interval, and can be defined as the $m^{\text{th}}$ derivative of a Legendre polynomial \cite[p.5]{belousov1962tables}: \nomenclature[Plm]{$\Plm$}{The Associated Legendre polynomial of degree $l$ and order $m$}
\bseqn{
	\Plm(x) &:= (-1)^m (1-x^2)^\frac{m}{2} \dmdxm{x}{m} P_l(x) = (-1)^m \; {(l+m)! \over 2^m \; l!} \; (1-x^2)^\frac{m}{2} \; P^{(m,m)}_{l-m}(x),
}
for $m = 0,1,2,\dots,l$. It is also standard to include a further definition for the case of $m$ negative -- that is we also define:
\bseqn{
	\Plm(x) &:= (-1)^m \; {(l+m)! \over (l-m)!} \; P_l^{\abs{m}}(x) = {(l+m)! \over 2^{\abs{m}} \; l!} \; (1-x^2)^{\abs{m} \over 2} \; \jaclmabs{l}{m}(x)
}
for $m = -l, \dots, -1$.

Once again, associated Legendre polynomials are special cases of \textit{associated Legendre functions} when $l$ is a non-negative integer and $m \in \{0,\dots,l\}$, which are also a class of univariate functions that are solutions to a modified version of the Legendre equation
\bseqn{
	(1 - x^2) \dmdxm{x}{2}y - 2x \ddx{x}y + \Big[l(l+1) - {m^2 \over 1-x^2} \Big] y = 0,
}
which is also helpfully known as the \textit{associated Legendre equation} (\cite[p.2]{belousov1962tables}, \cite[(14.2.2)]{DLMF}). We can see that Legendre functions (and polynomials) are just special cases of the associated versions when $m = 0$.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[(14.30.1)]{DLMF}: \nomenclature[Ylm]{$\Ylm$}{The degree $l$ and order $m$ spherical harmonic}
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where \nomenclature[clm]{$\clm$}{The normalising constant for the spherical harmonic $\Ylm$}
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0,
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0.
		\end{cases}
}
The spherical harmonics as defined are then orthonormal with respect to the inner product with uniform measure \nomenclature[conjuagate]{$a^*$}{The complex conjugate of $a \in \C$}
\bseqn{
	&\iint_{\Omega} \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \D S \\
	&\quad = \int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	&\quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	&\quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of \nomenclature[C]{$\C$}{The complex numbers} $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x$, $y$, and $z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x$, $y$, and $z$ for any \nomenclature[Z]{$\Z$}{The integers} $m \in \Z$. Indeed, they are polynomials in $x$, $y$, and $z$, which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in \nomenclature[R]{$\R$}{The real numbers} \nomenclature[Rd]{$\R^d$}{The d-dimensional coordinate space over the real numbers} $\R^3$.

@@

Jacobi operators are matrices that correspond to multiplication of the orthogonal polynomial basis, in this case the spherical harmonics, by our cartesian coordinates $x$, $y$, and $z$. We can obtain the entries to these operator matrices by determining the coefficients of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$ in terms of $Y^{m'}_{l'}(x,y,z)$ for any point $(x,y,z)$ on the unit sphere. Relations concerning the products of spherical harmonics have already been established, and so the expressions we desire here could just be seen as special cases of these. More specifically, the coefficients of the expansion in the SH basis of the product of two SHs are composed of Clebsch--Gordan coefficients \cite[p.231]{sakurai1995modern}. Clebsch--Gordan coefficients are important in quantum mechanics and angular momentum, with tables and algorithms existing to calculate them \cite{mcnamee1964tables}. By simply choosing appropriate combinations, one could gain the desired coefficients for the expansion of each of $x\,\Ylm(x,y,z)$, $y\,\Ylm(x,y,z)$, and $z\,\Ylm(x,y,z)$.

@@

	The first two results are simply consequences of the relationships established for incrementing and decrementing the parameters of Jacobi polynomials detailed in \cite[(18.9.5), (18.9.6)]{DLMF}, along with the symmetry relation that $P_{n}^{(\alpha, \beta)}(-z) \equiv (-1)^n P_{n}^{(\beta, \alpha)}(z)$. The final one is obtained by rearranging the three-term recurrence in \cite[(18.9.1), (18.9.2)]{DLMF}

@@

\remark The results in \bsreflemma{lemma:shxyzrelations} can be derived via the relationships established for products of spherical harmonics and Clebsch-Gordan coefficients as mentioned earlier \cite[p.231]{sakurai1995modern} but we include this proof so as to generalise to other OPs later on in later chapters.

@@

These recurrences lead to Jacobi operators that correspond to multiplication by $x$, $y$, and $z$ of the OPs. Define for $l \in \No$: \nomenclature[Pvector]{$\SHvec$}{Vector of the SH OPs} \nomenclature[Pvectorl]{$\SHvec_l$}{(Sub-)Vector of the degree $l$ SH OPs}
\bseqn{
	\SHvecl := 
		\begin{pmatrix}
			Y^{-l}_l \\
			\vdots \\
			Y^l_l
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\SHvec := 
		\begin{pmatrix}
			\SHvec_0 \\
			\hline
			\SHvec_1 \\
			\hline
			\SHvec_2 \\
			\vdots \\
		\end{pmatrix}.
}
Note that these OP vectors we have defined are grouped and ordered according to the degree $l$, and within each degree ordered by the Fourier mode $m$. Let the Jacobi matrices \nomenclature[J]{$J_x, J_y, J_z$}{Jacobi operator matrices} $J_x, J_y, J_z$ hence be given by 
\bseqnnumber{
	J_x \SHvec(x,y,z) = x \SHvec(x,y,z), \quad J_y \SHvec(x,y,z) = y \SHvec(x,y,z), \quad J_z \SHvec(x,y,z) = z \SHvec(x,y,z) \label{eqn:jacobimatsdef}
}
for $(x,y,z) \in \Omega$. While $J_x, J_y, J_z$ are not Jacobi matrices in the classical sense, they are block Jacobi matrices. However, it is useful to simply label them as such. These Jacobi matrices can act on the coefficients vector of a function's expansion in the spherical harmonic basis to implement the operation for multiplication by $x$, $y$ or $z$. For example, given sufficiently large $n \in \N$ let the function $f(x,y,z) : \Omega \to \C$ be given by its expansion \nomenclature[fvec]{$\vec{f}, \vec{\xi}$}{Lower case bold letters refer to vectors} \nomenclature[transpose]{$\vec{v}^\top, A^\top$}{Transpose of the vector $\vec{v}$ or matrix $A$ (non complex conjuagate)}
\bseqn{
	f(x,y,z) = \sum_{l=0}^{N} \sum_{m=-l}^{l} f_{l,m} \; \Ylm(x,y,z) = \SHvec(x,y,z)^\top \: \fvec
}
where 
\bseqn{
	\fvec_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			\vdots \\
			f_{l,-l}
		\end{pmatrix} \in \C^{2l+1}, 
	\quad \quad 
	\fvec := 
		\begin{pmatrix}
			\fvec_0 \\
			\vdots \\
			\fvec_N
		\end{pmatrix}.
}
We say that $\fvec$ is the \textit{coefficients vector} of the function $f$. Then $x f(x,y,z)$ is approximated by $\SHvec(x,y,z)^\top \: J_x^\top \: \fvec$. In other words, $J_x^\top \: \fvec$ is the coefficient vector for the expansion of the function $(x,y,z) \mapsto x \: f(x,y,z)$ in the spherical harmonics basis.

@@

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{spy-of-Jz}}
		\centering
		\caption{$J_z$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of the Jacobi operator matrices, showing their sparsity and structure, up to order $N=10$ (i.e. $(N+1)^2$ unknowns). Dots represent non-zero entries of the operator. Each are block-tridiagonal, with the blocks corresponding to polynomial degree.}
	\label{fig:sh-spy-jacobi}
\end{figure}

@@

In coefficient space, we wish to work only with vectors of coefficients for the expansion of a function, to which we can apply operator matrices to that represent differential or other operations. Naturally, we of course need a way to obtain the coefficients $f_{l,m}$. The coefficients can be calculated via the integral
\bseqn{
	f_{l,m} = \iint_\Omega f(\costheta \sinphi, \sintheta \sinphi, \cosphi) \: \Ylmfull^* \: \sinphi \: \D \varphi \: \D \theta
}\bstodo{the complex conj should be included here, shouldn't it?}
using the orthonormality of the spherical harmonics. As discussed in \bsrefchapter{CHAPTER:intro}, methods to calculate these coefficients exist known as spectral transforms, and are well established for the spherical harmonics (see e.g. \cite{mohlenkamp1999fast, suda2002fast}). Additionally, packages are available for public use that implement efficient spherical harmonics transforms, for example the FastTransforms package \cite{slevinsky2019fast}.

@@

Let $\vec{v}$ be the coefficients of the expansion of $v$ up to order $N$. The operator $V$ would then be the result of the following \textit{operator Clenshaw algorithm}:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \vec{0}, \: \vec{\xi}_{N+1} = \vec{0} \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \vec{v}_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(J_x,J_y,J_z)) - \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } V = V(x,y,z) \approx \xi_0 Y_0
}
where at each iteration, $\vec{\xi}_n$ is a vector of matrices\footnote{It may help to think of this symbolically. At each iteration, the entries of $\vec{\xi}_n$ are polynomials in $J_x$, $J_y$ and $J_z$. On the final step, $\vec{\xi}_{0} = \xi_{0}$ is hence some polynomial expression in $J_x$, $J_y$ and $J_z$, which we can then evaluate by inserting the Jacobi matrices.} (note that we are abusing notation here a bit, however it is the simplest way to present the algorithm without introducing yet more matrices!).
